---
output:
  word_document: default
---
# Paula McCree Bailey
## Module 2 - Assignment 1
### Correlation and Simple Linear Regression Assignment

```{r load library, message=FALSE}
##install.packages("lmtest")
##install.packages("GGally")
##install.packages("car")

library(tidyverse)
library(GGally)
library(car)
library(lmtest)

air = airquality
```


**TASK 1** Describe this dataset. How many variables and observations are there? Is there any missing data? Which variable is likely to be the response (Y) variable?

**The AirQuality dataset contains the daily air quality measurements in New York from May to September 1973.  It contains 6 variables with 153 observations. The dataset does contain missing data. The variable Ozone is likely to be the response (Y) variable.**


```{r Remove NA values}
air2 <- air %>% drop_na()
```

**TASK 2**  How many rows and columns remain in this new (air2) data frame?

**The new air2 dataframe contains 111 rows and 6 columns.**

```{r Visualization and calculate correlation}
#ggpairs plot for visualization and correlation

ggpairs(air2)
```


```{r create a correlation matrix}

#use "ggcorr" to create a correlation matrix with labels and correlation reported to two decimals

ggcorr(air2, label = "TRUE", label_round = 2)
```

**TASK 3**
Which variable is most strongly correlated with the “Ozone” variable?
Which variable is least strongly correlated with the “Ozone” variable?

**Temperature is more strongly correlated with the "ozone" variable.**
**Day is least strongly correlated with the "ozone" variable.**


```{r Graph temp vs ozone}

ggplot(air2,aes(x=Temp,y=Ozone)) + 
  geom_point()+
  labs(title = "Relationship Temperature vs Ozone")
```

**TASK 4**  Describe the relationship between “Temp” and “Ozone”.

**There seems to be a positive linear relationship between the temperature** **and ozone.  As the temperature increases, the ozone levels increase. The** **graph does have some outliers with higher ozone levels around 78 - 85** **degrees.**


```{r regression mode}
#create linear regression model

model1 <- lm(Ozone ~ Temp, air2)
summary(model1)
confint(model1)
```

**Task 5:** a. Discuss the quality of this model (mention the R square value and significance of the predictor variable). b. Use the code “confint(model1)” generate 95% confidence intervals for the coefficients. In what range does the slope coefficient likely fall?


**a.This seem to be an okay model. It seems odd that the Y-intercept **
**falls below the x-axis.  The slope is positive. For every 2.4391 degree ** **increase the ozone increases by 1.  Temp is significantly correlated with a** **p-value under .05.  The adjust R square is okay (not great), it is not**
**close to 1.**

**b. The slope is likely to fall in the range of [-0.47433, 5.35253].**




**TASK 6**  Re-do Task 4 to include the regression line

```{r graph w regression line}
ggplot(air2,aes(x=Temp,y=Ozone)) + 
  geom_point()+
  geom_smooth(method = "lm", color = "red", se = FALSE)+
  labs(title = "Temperature vs Ozone w/ Regression Line")
```

**TASK 7**  Develop a prediction for “Ozone” when “Temp” is 80.

```{r Question of linear relationship}

temp_pm <- filter(air2,air2$Temp==80)

#using predication function
predict(model1, newdata=temp_pm,interval="predict")

#prediction interval
temp_var <- predict(model1,interval = "prediction")
new_df <- cbind(air2, temp_var)

ggplot(new_df, aes(x = Ozone, y = Temp)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_line(aes(y=lwr), color = "red", linetype = "dashed") +
  geom_line(aes(y=upr), color = "red", linetype = "dashed") +
  theme_bw()
```

**Task 8** Perform appropriate model diagnostics to verify whether or not the model appears to meet the four linear regression model assumptions. 
Provide a brief comment on each assumptions validity for this model.

**The assumption was made the remaining portion of the assignment dealt**
**with the entire model and not just the subset of temps at 80 degrees.**

**Assumption 1: **The predictor and response variable have a linear relationship

**The predictor variable does seem to have a linear relationship with ozone. **
**The plot has many outliers, but it does seem to be linear.**


**Assumption 2:** Model errors (residuals) are independent

**We fail to reject the null hypothesis with a p-value greater than 0.05.** **This suggests that the residuals are likely independent.**

```{r Durbin-Watson Test}

dwtest(model1)
```

**Assumption 3: ** Model residuals exhibit constant variance  
Examine a plot of residuals.

```{r Add residuals}
#just temp values equal 80
#temp_pm = temp_pm %>% mutate(resid1 = model_80$residuals) #add the model residuals to our data frame
#ggplot(temp_pm,aes(x=Ozone,y=resid1)) + geom_point() + theme_bw()

# plot of residuals
air2 = air2 %>% mutate(resid1 = model1$residuals) #add the model residuals to our data frame
ggplot(air2,aes(x=Ozone,y=resid1)) + geom_point() + theme_bw()
```

**There does seem to appear to be a constant change in the variance of residuals. There are areas which are not constant, but overall it seems constant. **


**Assumption 4:**  Model residuals are Normally-distribute.
Examine a histogram of the residuals.

```{r graph - histogram}
#ggplot(temp_pm,aes(x=resid1)) + geom_histogram(bins = 15) + theme_bw()

ggplot(air2,aes(x=resid1)) + geom_histogram() + theme_bw()
```

```{r}
#qqPlot(model_80)

qqPlot(model1)
```
**The residuals histogram is Normal. The black line does not lie directly**
**in between the blue lines, but it is close. We also see outliers**

**The Normal Probability Plot below confirms what we saw in the histogram.**
**The data does follow a normal distribution.**



**Task 9 ** How might the model that you constructed in Task 5 be used?
Are there any cautions or concerns that you would have when recommending 
the model for use?

**The model might be used to predict the affects of temperature on the ozone.**

**Some cautions I would have is the outliers in the temperature.  For example**
**80 degrees produced variance in the ozone from 7 to 65.  This does not**
**seem likely to happen - maybe there was faulty equipment.**
